---
title: "Bad Neighbor BAP Development Notes"
date: "1-July-2018"
author: "Enrique Monta&ntilde;o, PhD"
output: html_notebook
---

The following outlines the various stages of development of the BAP.

## Testing

The main query source code is located in `"R/state_occurrence_query.R"`
The export processing code is located in `"R/export_bad_neighbor_list.R"`

The query requires two parameters
1) A single state FIPS code or a parenthetical group of FIPS codes.  Pairs of state FIPS codes and surrounding states are pre-developed and stored in `data/state_lookup.csv`.  Note: the list includes the Distict of Columbia;
2) A string hierarchy_homonym_string in the form: "*\\-179913\\-*".  Again, a predefined list is stored in: `data/heirarchy_strings.csv`.  The list of taxa is currently restricted to the best represented taxa in BISON.


```{r}
# compile the query query code
source("R/state_all_species_query.R")

# run the code for Virginia
# 51 FIPS code for Virginia
# list of surrounding state FIPS codes
buff_states <- "(24 11 37 47 21 54)"

# "*\\-202422\\-*" hierarchy_homonym_string for all plants
va_nn <- state_occurrence_query(fips_list = 51, taxon = "*\\-202422\\-*")

# view the result
# va_nn

buff_nn <- state_occurrence_query(fips_list = buff_states, taxon = "*\\-202422\\-*")

```

## Using lookup tables

We can now test the code using lookup tables.

First, the states:

```{r state-codes}
state_codes <- read_csv("data/state_lookup.csv")
state_codes
```

Next, the hierarchy homonym strings:

```{r taxon-strings}
taxa <- readr::read_csv("data/heirarchy_strings.csv")
taxa
```

Using the same example state above, Virgina, we can replicate the result using the lookup tables as input to the query.

```{r lookup-test}

va_data <- filter(state_codes, state_name == "Virginia")
va_data

va_nn_lookup <- state_occurrence_query(fips_list = va_data$state_fips, taxon = taxa$hierarchy_homonym_string[4])
va_nn_buffer_lookup <- state_occurrence_query(fips_list = va_data$buffer_fips, taxon = taxa$hierarchy_homonym_string[4])
```

We are getting the same result from both versions of the data, so that means the query using lookup works.

## Intersect results

The next step in the BAP is to find those species in the buffer states that are not current observed in the state of interest (Virgina in this example).  This is accomplished by intersecting the two lists.  The R `dplyr` library offers a convenient way to do this using various table joins.

```{r intersect-tables}
# Join the two data sets, excluding TSNs from Virgina
diff_species_buffer <- anti_join(va_nn_buffer_lookup, va_nn_lookup, by="tsn")
diff_species_buffer

```

There are 601 non-natives species documented in states bordering this state that are not documented in the state itself in BISON.  This is consistent with the results from the webservice: [NISC "Bad Neighbor" Analysis](http://www.stingersplace.com/js/BISONBadNeighbor.html), that runs the analysis based on ITIS scientific names instead of TSN.  The difference can be explain through the removal of ambiguous TSN (homonymns and synonyms) done in the query here.

Run the opposite join to confirm we get a different result for the species in Virginia that are not in the Buffer states.

```{r species-in-west}
# perform the opposite
diff_species_va <- anti_join(va_nn_lookup, va_nn_buffer_lookup, by="tsn")
diff_species_va
```

There is no direct comparison for this result, but it clarifies the results as being different and not related to the intersect itself.

## Export Results

The last step in the query is to export the results.  The export will run the intersect outlined above and result in a JSON dataset for portability and use in webservices.

```{r}

export_bad_neighbor_list(state_list = va_nn_lookup, buffer_list = va_nn_buffer_lookup, taxon = taxa$common_name[4], state_name = va_data$state_name)

# check the result
va_json <- jsonlite::fromJSON("result_json/Plants/Virginia_Plants_bad_neighbor.json")
# head(va_json)
```


The export works, so now we can put it all together to loop over all combinations of taxa and states.

## Generate all states

Loop over all combinations of states and taxa

```{r loop-all}
# load source scripts
source("R/export_bad_neighbor_list.R")
source("R/state_all_species_query.R")
source("R/state_bad_neighbor_query.R")

# load lookup tables
taxa <- readr::read_csv("data/heirarchy_strings.csv")
state_codes <- readr::read_csv("data/state_lookup.csv")

# a subset of the taxa to query
taxa_idx <- c(1,2,3,4,5,7,8,9)

# build a loop for the taxa
# for (k in 1:nrow(taxa)) {
for(k in taxa_idx) {
    # get the taxon
    taxon <- taxa[k, ]
   
    print(stringr::str_c("Processing Taxon:", taxon$common_name, sep = " "))
        
    # a list to hold results
    bad_neighbor_list <- vector("list", nrow(state_codes))

    # a list to hold the all-species result
    all_species_list <- vector("list", nrow(state_codes))

    # build a loop for all states
    for(i in 1:nrow(state_codes)) {

        print(stringr::str_c("Processing:", state_codes$state_name[i], "FIPS:", state_codes$state_fips[i],
                    "Buffer FIPS:", state_codes$buffer_fips[i], sep = " "))

        # run the occurrence queries
        state_nn <- state_bad_neighbor_query(fips_list = state_codes$state_fips[i], 
                                           taxon = taxon$hierarchy_homonym_string)
        
        buffer_nn <- state_bad_neighbor_query(fips_list = state_codes$buffer_fips[i], 
                                            taxon = taxon$hierarchy_homonym_string)

        # run the all-species query for the state
        all_species_list[[i]] <- state_all_species_query(fips_list = state_codes$buffer_fips[i], 
                                             taxon = taxon$hierarchy_homonym_string,
                                             state_name = state_codes$state_name[i])
        
        # test for an empty result
        if(!is.null(state_nn) && !is.null(buffer_nn)) {
            # export the results
            bad_neighbor_list[[i]] <- export_bad_neighbor_list(state_list = state_nn, buffer_list = buffer_nn, 
                                        taxon = taxon$common_name, state_name = state_codes$state_name[i])
        } 

    }

    # combine the results into a single tibble
    bad_neighbor_df <- dplyr::bind_rows(bad_neighbor_list)
    # combine the all-species
    all_species_df <- dplyr::bind_rows(all_species_list)

    # combine all the data frames
    all_results[[taxon$common_name]] <- list(bad_neighbor = bad_neighbor_df,
                                                      all_species = all_species_df)
}

# export the result
ar_json <- jsonlite::toJSON(all_results)
# write the json to file
readr::write_lines(ar_json, path = "result_json/all_taxa_summaries.json")

```

## Relative Contribution

Another part of the analysis considers the realitve abundance (or simply number as a count) of the non-native species present to all species in a given state.  This is simply the ratio of non-native to native species.  This is also easily plotted.  A new script will query BISON for the full list of species in a state and count the number of species and the abundance (number of records found).

Ideas for maps

1) as % of area
2) as % of all related taxa in state (e.g. ratio of nn plants to native plants)

## Relative Contribution

Another part of the analysis considers the realitve abundance (or simply number as a count) of the non-native species present to all species in a given state.  This is simply the ratio of non-native to native species.  This is also easily plotted.  A new script will query BISON for the full list of species in a state and count the number of species and the abundance (number of records found).


