---
title: "Bad Neighbor BAP Development Notes"
date: "Started: 1-July-2018"
author: "Enrique Monta&ntilde;o, PhD"
output: html_notebook
---

The following outlines the various stages of development of the BAP.

## Testing

The various functions for the BAP are as follows:

- `process_bad_neighbor_groups.R` is a wrapper function executing all necessary steps;
- `state_bad_neighbor_query.R` is the primary BISON query and parsing function to develop both the state and buffer state lists of non-native species;
- `state_all_species_query.R` is a second BISON query/parsing function returning all species in a given state.  This is included for possible future work, but the results are not used here;
- `export_bad_neighbor_list.R` is a wrapper function to run the intersect between states and the buffer states to create the bad neighbor list and export the results to JSON. Result files are located in the `results_json` directory;
- `generate_bn_count_map.R` is a function to generate a chloropleth map of results.


The main query requires two parameters
1) A single state FIPS code or a parenthetical group of FIPS codes.  Pairs of state FIPS codes and surrounding states are pre-developed and stored in `data/state_lookup.csv`.  Note: the list includes the District of Columbia;
2) A string hierarchy_homonym_string in the form: `"*\\-179913\\-*"`.  A predefined list is stored in: `data/hierarchy_strings.csv`, and several groups of related taxa are in files ending in `"*_hierarchy_strings.csv"`.  The list of taxa is currently restricted to the best represented taxa in BISON.


```{r query-test, message=FALSE, warning=FALSE}
# compile the query query code
source("R/state_bad_neighbor_query.R")

# run the code for Virginia
# 51 FIPS code for Virginia
# list of surrounding state FIPS codes
buff_states <- "(24 11 37 47 21 54)"

# "*\\-202422\\-*" hierarchy_homonym_string for all plants
va_nn <- state_bad_neighbor_query(fips_list = 51, taxon = "*\\-202422\\-*")

# run the query for the buffer states
buff_nn <- state_bad_neighbor_query(fips_list = buff_states, taxon = "*\\-202422\\-*")

# view the results
va_nn
buff_nn
```

We now have lists of non-native plant species for both the state of interest, Virginia, and the buffer states surrounding them.  The list is comprised of the scientific name of the species, the number of occurrences found, and the Taxonomic Serial Number (TSN) for the species from the Integrated Taxonomic Information System ([ITIS](https://www.itis.gov)).

## Using lookup tables

We can now test the code using lookup tables.

First, the states:

```{r state-codes, message=FALSE, warning=FALSE}
state_codes <- read_csv("data/state_lookup.csv")
state_codes
```

Next, the hierarchy homonym strings:

```{r taxon-strings, message=FALSE, warning=FALSE}
taxa <- readr::read_csv("data/hierarchy_strings.csv")
taxa
```

Using the same example state above, Virginia, we can replicate the all plants result using the lookup tables as input to the query.

```{r lookup-test, message=FALSE, warning=FALSE}
# filter the states list to VA
va_data <- filter(state_codes, state_name == "Virginia")
va_data

# run the queries
va_nn_lookup <- state_bad_neighbor_query(fips_list = va_data$state_fips, taxon = taxa$hierarchy_homonym_string[4])
va_nn_buffer_lookup <- state_bad_neighbor_query(fips_list = va_data$buffer_fips, taxon = taxa$hierarchy_homonym_string[4])

# View results
va_nn_lookup
va_nn_buffer_lookup
```

We are getting the same result from both versions of the data, so that means the query using lookup works.

## Intersect results

The next step in the BAP is to find those plant species in the buffer states that are not current observed in the state of interest (Virginia in this example); this is the bad neighbor list.  This is accomplished by intersecting the two lists.  The R `dplyr` library offers a convenient way to do this using various table joins.  The function uses the TSN for the intersect since the number should: 1) be unique to a species, and 2) not suffer from inconsistent spelling or string comparison error.

```{r intersect-tables, message=FALSE, warning=FALSE}
# Join the two data sets, excluding TSNs from Virgina
diff_species_buffer <- anti_join(va_nn_buffer_lookup, va_nn_lookup, by="tsn")

# the result is the list of bad neighbors
diff_species_buffer

```

There are 601 non-native plant species documented in states bordering this state that are not documented in the state itself in BISON.  Note that we are only interested in the number of species (number of rows in the result), not the count field.  The counts are recorded occurrences, but do not reflect a full distribution, so to report this number can be misleading.  However, reporting a documented species, i.e. a species with *any occurrence*, represents at least one known occurrence of the species.

The result is consistent with the results from the web service: [NISC "Bad Neighbor" Analysis](http://www.stingersplace.com/js/BISONBadNeighbor.html), that runs the analysis based on ITIS scientific names instead of TSN.  The difference can be explain through the removal of ambiguous TSN (homonyms and synonyms) done in the query here.

Run the opposite join to confirm we get a different result for the plant species in Virginia that are not in the Buffer states.

```{r species-in-va, message=FALSE, warning=FALSE}
# perform the opposite
diff_species_va <- anti_join(va_nn_lookup, va_nn_buffer_lookup, by="tsn")
diff_species_va
```

There is no direct comparison for this result, but it clarifies the results as being different and not related to the intersect itself.

## Export Results

The last step in the query is to export the results.  The export will run the intersect outlined above and result in a JSON data set for portability and use in web services.  The function `export_bad_neighbor_list` performs the intersect and exports the bad neighbor species list for each state.

```{r export-result-json, message=FALSE, warning=FALSE}
# run the export_bad_neighbor_list function to intersect and export the list
export_bad_neighbor_list(state_list = va_nn_lookup, buffer_list = va_nn_buffer_lookup, taxon = taxa$common_name[4], state_name = va_data$state_name)

# check the result
va_json <- jsonlite::fromJSON("result_json/Plants/Virginia_Plants_bad_neighbor.json")
head(va_json)
```


The export works, so now we can put it all together to loop over all combinations of taxa and states.

## Generate all states

Loop over all combinations of states and taxa.

```{r loop-all, eval=FALSE, message=FALSE, warning=FALSE}
# load source scripts
source("R/export_bad_neighbor_list.R")
source("R/state_all_species_query.R")
source("R/state_bad_neighbor_query.R")

# load lookup tables
taxa <- readr::read_csv("data/hierarchy_strings.csv")
state_codes <- readr::read_csv("data/state_lookup.csv")

# a subset of the taxa to query
taxa_idx <- c(1,2,3,4,5,7,8,9)

# build a loop for the taxa
# for (k in 1:nrow(taxa)) {
for(k in taxa_idx) {
    # get the taxon
    taxon <- taxa[k, ]
   
    print(stringr::str_c("Processing Taxon:", taxon$common_name, sep = " "))
        
    # a list to hold results
    bad_neighbor_list <- vector("list", nrow(state_codes))

    # a list to hold the all-species result
    all_species_list <- vector("list", nrow(state_codes))

    # build a loop for all states
    for(i in 1:nrow(state_codes)) {

        print(stringr::str_c("Processing:", state_codes$state_name[i], "FIPS:", state_codes$state_fips[i],
                    "Buffer FIPS:", state_codes$buffer_fips[i], sep = " "))

        # run the occurrence queries
        state_nn <- state_bad_neighbor_query(fips_list = state_codes$state_fips[i], 
                                           taxon = taxon$hierarchy_homonym_string)
        
        buffer_nn <- state_bad_neighbor_query(fips_list = state_codes$buffer_fips[i], 
                                            taxon = taxon$hierarchy_homonym_string)

        # run the all-species query for the state
        all_species_list[[i]] <- state_all_species_query(fips_list = state_codes$buffer_fips[i], 
                                             taxon = taxon$hierarchy_homonym_string,
                                             state_name = state_codes$state_name[i])
        
        # test for an empty result
        if(!is.null(state_nn) && !is.null(buffer_nn)) {
            # export the results
            bad_neighbor_list[[i]] <- export_bad_neighbor_list(state_list = state_nn, buffer_list = buffer_nn, 
                                        taxon = taxon$common_name, state_name = state_codes$state_name[i])
        } 

    }

    # combine the results into a single tibble
    bad_neighbor_df <- dplyr::bind_rows(bad_neighbor_list)
    # combine the all-species
    all_species_df <- dplyr::bind_rows(all_species_list)

    # combine all the data frames
    all_results[[taxon$common_name]] <- list(bad_neighbor = bad_neighbor_df,
                                                      all_species = all_species_df)
}

# export the result
ar_json <- jsonlite::toJSON(all_results)
# write the json to file
readr::write_lines(ar_json, path = "result_json/all_taxa_summaries.json")

```


