---
title: "R Notebook"
output: html_notebook
---

## Maps

```{r}
library(tidyverse)
source("R/generate_bn_count_map.R")
# ar <- jsonlite::fromJSON("result_json/all_taxa_summaries.json")

n <- names(all_results)
n 
# test with one of the results
df1 <- all_results[[n[3]]]$bad_neighbor

generate_bn_count_map(df1, taxon = n[3])

# create a loop for plotting all taxon
for (taxon in n) {
    print(stringr::str_c("Plotting: ", taxon))
    
    # create a file name
    out_name <- file.path("graphics", stringr::str_c("bad_neighbor_map_", taxon, ".png", sep = ""))
    # print(out_name)
    
    # generate the map
    m <- generate_bn_count_map(all_results[[taxon]]$bad_neighbor, taxon = taxon)
    
    # save
    ggplot2::ggsave(out_name, m, height = 6, width = 9)
}
  

```


Now we have plot of the actual count of species for each taxon, so we can visualize the threat to any given state.  However, we can also run some summary statistics based on cumulative counts for all taxa by state.  

- What states are under most threat?
- What contribution does each taxa make to the total?

We need to combine the various taxa dataframes into a single data frame with a conditional factor for taxon.  the final data frame should look something like:

state_name, taxon, species_count

```{r}
# Test
# comb_all_results <- dplyr::bind_rows("Birds" = all_results[[n[1]]]$bad_neighbor, 
                                     # "Plants" = all_results[[n[4]]]$bad_neighbor,
                                     # .id = "taxon")

# run for all
comb_all_results <- purrr::imap_dfr(all_results, ~ tibble::tibble(
    taxon = .y,
    state_name = .x$bad_neighbor$state_name,
    species_count = .x$bad_neighbor$species_count
))

comb_all_results

# summarize taxon results by state
comb_all_results %>% 
    # group by state then taxon
    group_by(state_name, taxon) %>% 
    summarise(species_count)


```


Pie chart of taxon threats

```{r}
# summarize the data on taxon
comb_all_results %>% 
    group_by(taxon) %>% 
    summarise(sum_species_count = sum(species_count)) %>%
    mutate(pct_species = (sum_species_count / sum(sum_species_count)) *100) %>% 
    ggplot(aes(x = taxon, y = pct_species)) +
    geom_bar(stat = "identity")


```



## Tests (NOT USED)
```{r}
us_states <- sf::st_as_sf(maps::map("state", plot = FALSE, fill = TRUE))
    
    # project to albers equal area for the U.S.
    # EPSG:102003 USA_Contiguous_Albers_Equal_Area_Conic
    # +proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs
us_states_alb <- sf::st_transform(us_states, crs = 102003)

ggplot2::ggplot() + ggplot2::geom_sf(data = us_states_alb)
    
    # mutate the state_name to match the map data
    df <- dplyr::mutate(df1, state_name = stringr::str_to_lower(stringr::str_replace_all(df1$state_name, "_", " ")))
    df
    # join the data
    us_states_2 <- dplyr::left_join(us_states_alb, df, by = c("ID" = "state_name"))
    
    us_states_2 <- dplyr::mutate(us_states_2, area = units::set_units(sf::st_area(us_states_alb), km^2))
    
    us_states_2$bn_density * 100
    
    
    
    # normalize the data based on area
    us_states_2 <- dplyr::mutate(us_states_2, bn_density = species_count / area)
    
    ggplot2::ggplot() + ggplot2::geom_sf(data = us_states_2, ggplot2::aes(fill = bn_density * 100)) #+
        ggplot2::scale_fill_gradient(name = "Bad Neighbors", low = "#fff7fb", high = "#014636")
    
    # build a title, replacing _ as needed
    t <- stringr::str_c("Lower 48", stringr::str_replace_all(taxon, "_", " "), 
                        "Bad Neighbor Threat Density", sep = " ") 

    
    
    https://bison.usgs.gov/solr/occurrences/select/?q=computedStateFips:51%20AND%20hierarchy_homonym_string%3A%2A%5C-202422%5C-%2A%20&establishmentMeans:L48&rows=0&facet=true&facet.field=ITISscientificName&facet.limit=-1&facet.mincount=1&wt=json
```

