---
title: "Bad Neighbor Analysis Biogeographic Analysis Package"
author: "Enrique L. Monta&ntilde;o, PhD"
date: "Started: July-1-2018"
output: html_notebook
---

```{r libraries, message=FALSE, warning=FALSE, results='hide'}
# libraries for the notebook
library(tidyverse)
source("R/generate_bn_count_map.R")
```


## Summary

There are approximately 6000 distinct non-native species (plant and animal) currently documented in the Lower 48 United States.  Not every species is found or will be found in a given state.  The purpose of this provisional bio-geographic analysis package is to create a state-by-state list of regulated non-native (i.e. invasive) species found in adjacent states, but not yet documented in the state of interest.  Stakeholders in states can use the "Bad Neighbor" analysis to see a list of species of concern to develop monitoring and management strategies before invasion.

The Bad Neighbor analysis can be run on any type of taxon for which the Biodiversity Information Serving Our Nation (BISON) species occurrence database has records.  For the sake of illustration, a subset of taxonomic classes and families were selected to demonstrate the analysis.  Those groups include species of high concern in the plant, animal and insect kingdoms.  Groups chosen were:

1) Specific groups of plants:  
    - Tree/Shrub  
    - Grass  
    - Ivy/Vines  
    - Forb/Herb  
2) Insects  
3) Birds  


## Processing steps

The following outlines the various stages of running the Bad Neighbor BAP.  The process is developed in three primary steps:

State list  
1) Query individual state for a list of all species at a given taxonomic group.  The species group list can be generated from any valid taxonomic group;

Buffer States List  
2) Query adjacent states for a list of all species at a given taxonomic group;

Bad Neighbor List  
3) Intersect lists 1 and 2 to find a list of species in 2 (buffer states) not currently documented in 1 (state); this is the Bad Neighbor list.

The processing is done in several R scripts stored in the `R` directory:

- `process_bad_neighbor_groups.R` is a wrapper function executing all necessary steps;
- `state_bad_neighbor_query.R` is the primary BISON query and parsing function to develop both the state and buffer state lists of non-native species;
- `state_all_species_query.R` is a second BISON query/parsing function returning all species in a given state.  This is included for possible future work, but the results are not used here;
- `export_bad_neighbor_list.R` is a wrapper function to run the intersect between states and the buffer states to create the bad neighbor list and export the results to JSON. Result files are located in the `results_json` directory;
- `generate_bn_count_map.R` is a function to generate a chloropleth map of results.


## Results

The results have been run and saved in a csv file.

```{r load-combined-results, message=FALSE, warning=FALSE}
comb_all_results <- readr::read_csv("result_csv/combine_all_plant_bad_neighbor.csv")
```

Of the 6000 non-native species, there are a total of `r sum(comb_all_results$bad_neighbor_count)` "Bad Neighbor" occurrences of those species in the Lower 48 United States.  The table below describes the results by taxon group as total and the relative contribution, or threat, for each group.

```{r summary-results-table, echo=FALSE}
comb_all_results %>% 
    group_by(family) %>% # group by family
    summarise(contrib_bad_neighbor = sum(bad_neighbor_count)) %>%  # calculate the sum of each group
    mutate(pct_contrib_bad_neighbor = (contrib_bad_neighbor / sum(contrib_bad_neighbor) * 100)) %>% 
    arrange(pct_contrib_bad_neighbor) %>% 
    rename(`Taxon Group` = family,
           `Number of Bad Neighbors` = contrib_bad_neighbor,
           `Pct of Bad Neighbor` = pct_contrib_bad_neighbor)
```

The chart below illustrates the relative threat from each of the taxon groups.

```{r summary-results-chart, echo=FALSE}
comb_all_results %>% 
    group_by(family) %>% # group by family
    summarise(contrib_bad_neighbor = sum(bad_neighbor_count)) %>%  # calculate the sum of each group
    mutate(pct_contrib_bad_neighbor = (contrib_bad_neighbor / sum(contrib_bad_neighbor) * 100)) %>% 
    ggplot(aes(x = reorder(family, pct_contrib_bad_neighbor), y = pct_contrib_bad_neighbor)) +
    geom_bar(stat = "identity") +
    # geom_text(aes(label = paste0(prettyNum(pct_contrib_bad_neighbor, digits = 3), "%"), 
                  # y = pct_contrib_bad_neighbor),
                  # vjust = 1.4, size = 4, color = "white") +
    labs(title = "Threat of Plant Family Bad Neighbors",
         x = "Plant Family Bad Neighbor",
         y = "Percent Contribution") +
    # ggthemes::theme_tufte()
    theme_bw() +
    theme(plot.title = ggplot2::element_text(size = ggplot2::rel(1.5)))
```

The results indicate the Forb/Herb group present the greatest overall threat, with Tree/Shrub, Insects, and Grass close behind.  Recall, a Bad Neighbor is defined as a species that is not yet documented in a given state.  Therefore, these results suggest many states are not yet invaded.  Birds are a smaller Bad Neighbor, possibly because of migration and ease of movement.  The low (almost zero) results for Ivy/Vines suggest the species in this group *already* occur in the states.


## Maps

Maps can help describe the problem and where it is occurring.  A choropleth map describes the location of the states under greatest threat.  The results of the Bad Neighbor Analysis contain the number of species in adjacent states that are not currently documented in the state of interest.  Therefore, plotting this count as a color intensity describes those states that are under most threat.  The first map is for all groups combined.


```{r threat-map, echo=FALSE}
comb_all_results %>% 
    group_by(state_name) %>% 
    summarise(species_count = sum(bad_neighbor_count)) %>% 
    generate_bn_count_map(taxon = "All Groups")

```

The summary map shows all states have Bad Neighbors from some group.  The areas of red in the map indicate those states under the greatest threat from its neighbors; that is, there are an increasing number of non-native species in neighboring states that are not yet documented in the state itself.  The package provides a list of the actual bad neighbor species a land manager in the state might use to prioritize monitoring.

States that are colored green present a more nuanced result.  Most simply, a lower number of bad neighbors can indicate a lack of non-native species.  However, a more realistic interpretation is the state is already host to most of the non-native species present in adjacent states.  Florida, for example, is dark green and only has 171 Bad Neighbors, but is known to already be host to many non-native species.


### Taxon Group Results

The result maps from each individual taxon group are presented below in descending order.

**Forb/Herb**

```{r forb-herb-map, echo=FALSE, message=FALSE, warning=FALSE}
comb_all_results_forbs <- readr::read_csv("result_csv/forbs_comb_result.csv")

# summarize the results to count all tree_shrub species per state
comb_all_results_forbs %>% 
    group_by(state_name) %>% 
    summarize(species_count = sum(bad_neighbor_count)) %>% 
    # pass to the plot function
    generate_bn_count_map(taxon = "Forb/Herb")
```

**Tree/Shrub**

```{r tree-shrib-map, echo=FALSE, message=FALSE, warning=FALSE}
comb_all_results_trees <- readr::read_csv("result_csv/trees_comb_result.csv")

# summarize the results to count all tree_shrub species per state
comb_all_results_trees %>% 
    group_by(state_name) %>% 
    summarize(species_count = sum(bad_neighbor_count)) %>% 
    # pass to the plot function
    generate_bn_count_map(taxon = "Tree/Shrub")
```

**Insect**

```{r insect-map, echo=FALSE, message=FALSE, warning=FALSE}
# load the saved file
comb_all_results_insects <- readr::read_csv("result_csv/insects_comb_result.csv")


# summarize the results to count all tree_shrub species per state
comb_all_results_insects %>% 
    group_by(state_name) %>% 
    summarize(species_count = sum(bad_neighbor_count)) %>% 
    # pass to the plot function
    generate_bn_count_map(taxon = "Insect")
```

**Grass**

```{r grass-map, echo=FALSE, message=FALSE, warning=FALSE}
# load the saved file
comb_all_results_grasses <- readr::read_csv("result_csv/grasses_comb_result.csv")

# summarize the results to count all tree_shrub species per state
comb_all_results_grasses %>% 
    group_by(state_name) %>% 
    summarize(species_count = sum(bad_neighbor_count)) %>% 
    # pass to the plot function
    generate_bn_count_map(taxon = "Grass")
```


**Birds**

```{r bird-map, echo=FALSE, message=FALSE, warning=FALSE}
# load the saved file
comb_all_results_birds <- readr::read_csv("result_csv/birds_comb_result.csv")

# summarize the results to count all tree_shrub species per state
comb_all_results_birds %>% 
    group_by(state_name) %>% 
    summarize(species_count = sum(bad_neighbor_count)) %>% 
    # pass to the plot function
    generate_bn_count_map(taxon = "Bird")
```


**Ivy/Vine**

```{r ivy-map, echo=FALSE, message=FALSE, warning=FALSE}
# load the saved file
comb_all_results_ivy <- readr::read_csv("result_csv/ivy_comb_result.csv")

# summarize the results to count all tree_shrub species per state
comb_all_results_ivy %>% 
    group_by(state_name) %>% 
    summarize(species_count = sum(bad_neighbor_count)) %>% 
    # pass to the plot function
    generate_bn_count_map(taxon = "Ivy/Vine")
```


### Demonstration of state-level results

The following is an illustration of how specific taxon group results might be presented to a user when a particular state is selected.  The user would "click", or select, a state of interest, and be presented with a summary of the results and list of Bad Neighbor species in the taxon group.

The `generate_state_report.R` function was written to produce the text:

```{r demo-report, warning=FALSE, message=FALSE}
source("R/generate_state_report_text.R")

# print a text report
cat(generate_state_report_text("Colorado", group_name = "Insect", comb_all_results_insects))
```

The species list is stored as json in the `results_json` directory.  The data were generated on a taxon group-by-group basis, then by state.  The separate taxon group json files (e.g. tree/shrub, ivy/vines, forb/herb, grasses) need to be combined into a single list for plant groups.  The insect group will be used for illustration since this is only a single taxon group.

The final result can combine the list of scientific names with a link to BISON for more information.

```{r demo-table, warning=FALSE, message=FALSE}
# load the insect result file.  This includes all summaries and species list
species_list <- jsonlite::fromJSON("result_json/insect/Insecta/Colorado_Insecta_bad_neighbor.json")

# parse the result to generate a table of species
species_list$species_list %>% 
    select(name, tsn) %>% 
    # create an html link from the result
    mutate(link = str_c("<a href ='https://bison.usgs.gov/index.jsp?scientificName=",
                        name, 
                        "&ITIS=itis' target='_blank'>")) %>% 
    rename("ScientificName" = name, "TSN" = tsn) #rename for clarity
```



