---
title: "Development of taxon group results"
output: html_notebook
---

```{r}
library(tidyverse)
source("R/export_bad_neighbor_list.R")
source("R/state_all_species_query.R")
source("R/state_bad_neighbor_query.R")
source("R/process_bad_neighbor_groups.R")
source("R/generate_bn_count_map.R")
```

Process:

State list
1) Query individual state for a list of all species at a given taxonomic Class.  The Class list can be generated from ranks within the non-native species list with the understanding all Classes can be run;
2) Intersect the results from 1 with the non-native list to find the non-native species currently found in the state;

Buffer List
3) Query adjacent states for a list of all species at a given taxonomic Class;
4) Intersect the results from 3 with the non-native list to find the non-native species currently found in bordering states;

Bad Neighbor List
5) Intersect lists 2 and 4 to find a list of species in 4 (buffer states) not currently documented in 2 (state); this is the Bad Neighbor list.


This portion of the develpment uses Annie's non-native species list to verify and qualify results from BISON.  The non-native list is not yet published, and will not be used in the final product.  However, the BISON query uses these data to indicate non-native status.

### List of taxonomic Classes

```{r review-taxonomy}
# load the non-native master list from Annie
nn_list <- readxl::read_excel("data/Non-native_Master_07_06_2018.xlsx")

# reduce list to just L48 region, remove NA and undefined
# there are 690 rows without taxonomy, so drop those for now
nn_list <- nn_list %>% 
    filter(Region == "L48" & !is.na(Class) & Class != "undefined")

# look at the result
head(nn_list)

# find the distinct Class and count the number
class_list <- nn_list %>% 
    group_by(Class) %>% 
    summarize(num_species = n()) %>% 
    arrange(desc(num_species)) 
# print the list
class_list

nn_list %>% 
    group_by(Family) %>% 
    summarize(num_species = n()) %>% 
    arrange(desc(num_species)) 
    

```

Taxonomy does not accurately describe how an end user might approach invasive species.  For example, at the Class level, there are 56 entries describing the non-native species.  A more intuitive approach might be to group species into informal classifications like tree/shrub.  The problem here is there is no one Family or Class that describes this sub-group; hence informal.  The USDA PLANTS database provides a habitat flag to describe the generalized growth habit of various species like tree, shrub, subshrub, or combinations thereof.  Therefore, a list of Families that incorporate tree/shrub can be developed from the PLANTS habitat.

A series of families related by habit have been formed and extracted from the non-native list.  A bad neighbor analysis is run for each of these groups.

### Tree/shrub Families

```{r determine-trees}
nn_list %>% 
    filter(Family %in% c("Myrtaceae", "Fabaceae", "Ulmaceae", "Rosaceae")) %>% 
    group_by(Family) %>% 
    summarize(num_species = n()) %>% 
    arrange(desc(num_species)) 
```

The following runs the analysis using the original, BISON-only query (uses establishmentMeans:L48).  The results are similar using the nn-native list, and using BISON only is reproduceable without needing external lists.

```{r query-trees}

# load lookup tables
taxa <- readr::read_csv("data/tree_shrub_heirarchy_strings.csv")

trees_results <- process_bad_neighbor_groups("tree_shrub", taxa = taxa)

```

Now we have results for just tree/shrubs. From this result we can use those maps to produce a visualized threat assessment.

```{r extract-trees}
# vector of tree/shrub families
# trees <- c("Myrtaceae", "Fabaceae", "Ulmaceae", "Rosaceae")

# comb_all_results_trees <- purrr::imap_dfr(trees_results, ~ tibble::tibble(
#     taxon = .y,
#     state_name = .x$bad_neighbor$state_name,
#     bad_neighbor_count = .x$bad_neighbor$bad_neighbor_count,
#     state_nn_species_count = .x$bad_neighbor$state_nn_species_count
# ))
# # save the output
# readr::write_csv(comb_all_results_trees, "result_csv/trees_comb_result.csv")

comb_all_results_trees <- readr::read_csv("result_csv/trees_comb_result.csv")

# summarize the results to count all tree_shrub species per state
comb_all_results_trees %>% 
    group_by(state_name) %>% 
    summarize(species_count = sum(bad_neighbor_count)) %>% 
    # pass to the plot function
    generate_bn_count_map(taxon = "Tree/Shrub")

# save the output
# out_name <- file.path("graphics", stringr::str_c("bad_neighbor_map_", "tree_shrub", ".png", sep = ""))
# ggsave(out_name, height = 6, width = 9)

# print a text report
cat(generate_state_report_text("Alabama", group_name = "Tree/Shrub", comb_all_results_trees))

```

Now process the other types

### Forb/Herbs

```{r determine-forbs}
nn_list %>% 
    filter(Family %in% c("Asteraceae", "Pteridaceae", "Onagraceae", "Lamiaceae")) %>% 
    group_by(Family) %>% 
    summarize(num_species = n()) %>% 
    arrange(desc(num_species)) 
```

```{r process-forbs}
# load lookup tables
# taxa <- readr::read_csv("data/forb_herb_heirarchy_strings.csv")

# forbs_results <- process_bad_neighbor_groups("forbs_herbs", taxa = taxa)
# 
# comb_all_results_forbs <- purrr::imap_dfr(forbs_results, ~ tibble::tibble(
#     taxon = .y,
#     state_name = .x$bad_neighbor$state_name,
#     bad_neighbor_count = .x$bad_neighbor$bad_neighbor_count,
#     state_nn_species_count = .x$bad_neighbor$state_nn_species_count
# ))
# 
# # save the result
# readr::write_csv(comb_all_results_forbs, "result_csv/forbs_comb_result.csv")

# load the saved file
comb_all_results_forbs <- readr::read_csv("result_csv/forbs_comb_result.csv")


# summarize the results to count all tree_shrub species per state
comb_all_results_forbs %>% 
    group_by(state_name) %>% 
    summarize(species_count = sum(bad_neighbor_count)) %>% 
    # pass to the plot function
    generate_bn_count_map(taxon = "Forb/Herb")

# save the output
# out_name <- file.path("graphics", stringr::str_c("bad_neighbor_map_", "forb_herbs", ".png", sep = ""))
# ggsave(out_name, height = 6, width = 9)

```


### Vines/ivy

```{r}
nn_list %>% 
    filter(Family %in% c("Araceae", "Anacardiaceae", "Vitaceae")) %>% 
    group_by(Family) %>% 
    summarize(num_species = n()) %>% 
    arrange(desc(num_species)) 
```

```{r process-vines}
# load lookup tables
# taxa <- readr::read_csv("data/ivy_vines_heirarchy_strings.csv")
# 
# ivy_results <- process_bad_neighbor_groups("ivy_vines", taxa = taxa)
# 
# comb_all_results_ivy <- purrr::imap_dfr(ivy_results, ~ tibble::tibble(
#     taxon = .y,
#     state_name = .x$bad_neighbor$state_name,
#     bad_neighbor_count = .x$bad_neighbor$bad_neighbor_count,
#     state_nn_species_count = .x$bad_neighbor$state_nn_species_count
# ))
# save the result
# readr::write_csv(comb_all_results_ivy, "result_csv/ivy_comb_result.csv")

# load the saved file
comb_all_results_ivy <- readr::read_csv("result_csv/ivy_comb_result.csv")

# summarize the results to count all tree_shrub species per state
comb_all_results_ivy %>% 
    group_by(state_name) %>% 
    summarize(species_count = sum(bad_neighbor_count)) %>% 
    # pass to the plot function
    generate_bn_count_map(taxon = "Ivy/Vine")

# save the output
# out_name <- file.path("graphics", stringr::str_c("bad_neighbor_map_", "ivy_vines", ".png", sep = ""))
# ggsave(out_name, height = 6, width = 9)

```


### Grasses

```{r}
nn_list %>% 
    filter(Family %in% c("Cyperaceae", "Poaceae")) %>% 
    group_by(Family) %>% 
    summarize(num_species = n()) %>% 
    arrange(desc(num_species)) 
```

```{r process-grass}
# load lookup tables
# taxa <- readr::read_csv("data/grasses_heirarchy_strings.csv")
# 
# grasses_results <- process_bad_neighbor_groups("grasses", taxa = taxa)
# 
# comb_all_results_grasses <- purrr::imap_dfr(grasses_results, ~ tibble::tibble(
#     taxon = .y,
#     state_name = .x$bad_neighbor$state_name,
#     bad_neighbor_count = .x$bad_neighbor$bad_neighbor_count,
#     state_nn_species_count = .x$bad_neighbor$state_nn_species_count
# ))
# 
# # save the result
# readr::write_csv(comb_all_results_grasses, "result_csv/grasses_comb_result.csv")

# load the saved file
comb_all_results_grasses <- readr::read_csv("result_csv/grasses_comb_result.csv")


# summarize the results to count all tree_shrub species per state
comb_all_results_grasses %>% 
    group_by(state_name) %>% 
    summarize(species_count = sum(bad_neighbor_count)) %>% 
    # pass to the plot function
    generate_bn_count_map(taxon = "Grass")

# save the output
# out_name <- file.path("graphics", stringr::str_c("bad_neighbor_map_", "grasses", ".png", sep = ""))
# ggsave(out_name, height = 6, width = 9)

```

### Birds

```{r proc-birds}
# taxa <- readr::read_csv("data/heirarchy_strings.csv")

# process the insect taxa, record 1
# birds_results <- process_bad_neighbor_groups("bird", taxa = taxa[1,])
# # 
# comb_all_results_birds <- purrr::imap_dfr(birds_results, ~ tibble::tibble(
#     taxon = .y,
#     state_name = .x$bad_neighbor$state_name,
#     bad_neighbor_count = .x$bad_neighbor$bad_neighbor_count,
#     state_nn_species_count = .x$bad_neighbor$state_nn_species_count
# ))
# 
# # save the result
# readr::write_csv(comb_all_results_birds, "result_csv/birds_comb_result.csv")

# load the saved file
comb_all_results_birds <- readr::read_csv("result_csv/birds_comb_result.csv")


# summarize the results to count all tree_shrub species per state
comb_all_results_birds %>% 
    group_by(state_name) %>% 
    summarize(species_count = sum(bad_neighbor_count)) %>% 
    # pass to the plot function
    generate_bn_count_map(taxon = "Bird")

# save the output
# out_name <- file.path("graphics", stringr::str_c("bad_neighbor_map_", "birds", ".png", sep = ""))
# ggsave(out_name, height = 6, width = 9)
```



### Insects

```{r proc-insects}
# load lookup tables
# taxa <- readr::read_csv("data/heirarchy_strings.csv")
# 
# # process the insect taxa, record 3 
# insects_results <- process_bad_neighbor_groups("insect", taxa = taxa[3,])
# # 
# comb_all_results_insects <- purrr::imap_dfr(insects_results, ~ tibble::tibble(
#     taxon = .y,
#     state_name = .x$bad_neighbor$state_name,
#     bad_neighbor_count = .x$bad_neighbor$bad_neighbor_count,
#     state_nn_species_count = .x$bad_neighbor$state_nn_species_count
# ))
# 
# # save the result
# readr::write_csv(comb_all_results_insects, "result_csv/insects_comb_result.csv")

# load the saved file
comb_all_results_insects <- readr::read_csv("result_csv/insects_comb_result.csv")


# summarize the results to count all tree_shrub species per state
comb_all_results_insects %>% 
    group_by(state_name) %>% 
    summarize(species_count = sum(bad_neighbor_count)) %>% 
    # pass to the plot function
    generate_bn_count_map(taxon = "Insect")

# save the output
# out_name <- file.path("graphics", stringr::str_c("bad_neighbor_map_", "insects", ".png", sep = ""))
# ggsave(out_name, height = 6, width = 9)
```


## Summaries

What contribution does each group make to the total of bad neighbor plants?  

```{r summary-stats}
# trees
# sum_trees <- comb_all_results_trees %>%
#     group_by(state_name) %>%
#     summarize(bad_neighbor_count = sum(bad_neighbor_count), state_nn_species_count = sum(state_nn_species_count)) %>%
#     mutate(family = "Tree/Shrub")
# 
# # forbs
# sum_forbs <- comb_all_results_forbs %>%
#     group_by(state_name) %>%
#     summarize(bad_neighbor_count = sum(bad_neighbor_count), state_nn_species_count = sum(state_nn_species_count)) %>%
#     mutate(family = "Forb/Herb")
# 
# # grasses
# sum_grasses <- comb_all_results_grasses %>%
#     group_by(state_name) %>%
#     summarize(bad_neighbor_count = sum(bad_neighbor_count), state_nn_species_count = sum(state_nn_species_count)) %>%
#     mutate(family = "Grass")
# 
# # vines
# sum_ivy <- comb_all_results_ivy %>%
#     group_by(state_name) %>%
#     summarize(bad_neighbor_count = sum(bad_neighbor_count), state_nn_species_count = sum(state_nn_species_count)) %>%
#     mutate(family = "Ivy/Vine")
# 
# # birds
# sum_birds <- comb_all_results_birds %>%
#     group_by(state_name) %>%
#     summarize(bad_neighbor_count = sum(bad_neighbor_count), state_nn_species_count = sum(state_nn_species_count)) %>%
#     mutate(family = "Bird")
# 
# # insects
# sum_insects <- comb_all_results_insects %>%
#     group_by(state_name) %>%
#     summarize(bad_neighbor_count = sum(bad_neighbor_count), state_nn_species_count = sum(state_nn_species_count)) %>%
#     mutate(family = "Insect")
# 
# 
# # combine all
# comb_all_results <- bind_rows(sum_trees, sum_ivy, sum_forbs, sum_grasses, sum_birds, sum_insects)
# 
# # save the results
# readr::write_csv(comb_all_results, "result_csv/combine_all_plant_bad_neighbor.csv")

# load the results
comb_all_results <- readr::read_csv("result_csv/combine_all_plant_bad_neighbor.csv")

comb_all_results %>% 
    group_by(family) %>% # group by family
    summarise(contrib_bad_neighbor = sum(bad_neighbor_count)) %>%  # calculate the sum of each group
    mutate(pct_contrib_bad_neighbor = (contrib_bad_neighbor / sum(contrib_bad_neighbor) * 100)) %>% 
    ggplot(aes(x = reorder(family, pct_contrib_bad_neighbor), y = pct_contrib_bad_neighbor)) +
    geom_bar(stat = "identity") +
    # geom_text(aes(label = paste0(pct_contrib_bad_neighbor, "%"), y = pct_contrib_bad_neighbor),
            # vjust = 1.4, size = 5, color = "white") +
    labs(title = "Threat of Plant Family Bad Neighbors",
         x = "Plant Family Bad Neighbor",
         y = "Percent Contribution") +
    # ggthemes::theme_tufte()
    theme_bw() +
    theme(plot.title = ggplot2::element_text(size = ggplot2::rel(1.5)))
    
```




